---
title: "Class 9: Candy Mini Project"
author: "Andrew Duong (PID: A18129684)"
format: pdf
toc: true
---

## Background

In today's mini project we will analyze candy data with exploratory graphics, basic statistics, correlation analysis and principal component analusis methods we have been learning.

## Data Import

The data comes as a CSV file from 538. 

```{r}
candy_url <- url("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv")

candy = read.csv(candy_url, row.names = 1)
head(candy)
```

## What is in the dataset?

> Q1. How many different candy types are in this dataset?
There are `r nrow(candy)` rows in this dataset

```{r}
nrow(candy)
```


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruit candy types in the dataset.

## What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Skittles original",]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

## Side-Note: the skimr::skim() function

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The "hist" column seems to be on a different scale to the majority of the other columns.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero likely represents that there is no chocolate and one means the candy contains chocolate.

## Exploratory Analysis

> Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins = 20, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Win Percentages", 
       x = "Win Percentage", y = "Count")
```


> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50%. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Steps to solve this:
1. Find all chocolate candy in the dataset
2. Extract or find their winpercent values
3. Calculate the mean of these values

4. Find all fruity candy
5. Find their winpercent values
6. Calculate their mean value

```{r}
choc.candy <- candy[ candy$chocolate==1,]
choc.win <- choc.candy$winpercent
mean(choc.win)


```
```{r}
fruity.candy <- candy[ candy$fruity==1,]
fruity.win <- fruity.candy$winpercent
mean(fruity.win)
```


The mean for chocolate is higher than fruity. 

> Q12. Is this difference statistically significant?

The difference seems to be statistically significant as the p-value is below 0.5.

```{r}
t.test(choc.win, fruity.win)
```

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

The least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy[order(candy$winpercent, decreasing = T)[1:5], ]
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(x = winpercent, y = rownames(candy)) +
  geom_col() +
  ylab("") # turn off Y-label that we don't need

ggsave("barplot1.png", height = 40, width= 6)
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() +
  ylab("") # turn off Y-label that we don't need

ggsave("barplot2.png", height = 10, width= 6)
```


## Time to add some useful color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Taking a look at pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(pricepercent, winpercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The Reese's Minatures is the highest ranked in terms of winpercent for the least money.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

They are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate.

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar are most postively correlated.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)

```
```{r}
plot(pca$x[,1:2])

```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)

```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
  aes(x=PC1, y=PC2,
      size=winpercent/100,
      text=rownames(my_data),
      label=rownames(my_data)) +
  geom_point(col=my_cols)

p

```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}

loadings <- data.frame(pca$rotation)
loadings$variable <- rownames(loadings)

ggplot(loadings) +
  aes(x = reorder(variable, PC1), y = PC1) +
  geom_col() +
  coord_flip() 


```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The variables that are strongly positive on PC1 are fruity, pluribus, and hard which makes sense. We saw this relationship previously in the correlation matrix, winpercent comparisons, and bar rankings.

## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A winning candy tends to be candy bars, chocolate candies, contains peanuts or caramel, priced moderately, and high in sugar. The visualization shows that chocolate bars rank high. Correlation shows that chocolate and bars are postively correlated. PCA shows that they separate chocolate bars from fruity candies.