---
title: "Class 8: Analysis of Breast Cancer Cells"
author: "Andrew Duong: (PID: A18129684)"
format: pdf
toc: true
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data. 

## Data Import

The data is in CSV format:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)

```

wee peak at the data
```{r}
head(wisc.df, 3)
```

```{r}
nrow(wisc.df)
```

> Q1. How many observations are in this dataset?

There are 569 observations in the dataset.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis == "M")
```
There are 212 individuals who are malignant in their biopsy sites.

> Q3. How many variables/features in the data are suffixed with _mean?

There are 10 variables that are suffixed with _mean.

```{r}
length(grep("_mean", colnames(wisc.df)))
```

We need to remove the `diagnosis` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a separate wee vector that we can use later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis

The main function in base R is called `prcomp()`we will use the optional argument `scale=TRUE` here as the data columns/features/dimensions are on very different scales in the original data set.

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```

```{r}
wisc.pr$x
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col = diagnosis) + 
  geom_point()
```

```{r}
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

There need to at least be three PCS to describe 70% of the original variance.


> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

There are 7 PCS required to describe 90% of the original variance in the data.

## Interpreting PCA Results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is difficult to understand because distinctions cannot be easily made. 

## Variance Explained
```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var / sum(pr.var)

plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )


```



> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation[,1]
```

## Hierarchical Clustering

The goal of this section is to do hierarchal clustering of the original data. Recall from class that this type of clustering does not assume in advance the number 

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```

## Selecting numbers of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```
## Using different methods


```{r}
wisc.hclust.ward <- hclust(data.dist, method = "ward.D2")
wisc.hclust.clusters.ward <- cutree(wisc.hclust.ward, k = 4)
table(wisc.hclust.clusters.ward, diagnosis)
```



## Combining methods

The idea here is that I can take my new variables (i.e. the scores on the PCS `wisc.pr$x`) that are better descriptors of the data-set than the original features (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering. 

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k = 2 )
table(diagnosis)
```
```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```


I can now run `table()` with both my clustering `grps` and the expert `diagnosis`

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
table(grps, diagnosis)
```
> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

wisc.hclust.clusters seems to be better in separating the diagnoses. 

```{r}
table(wisc.hclust.clusters, diagnosis)
```
```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```


Our cluster "1" has 179 "M" diagnosis 
or cluster "2" has 333 "B" diagnosis

179 TP
24  FP

Sensitivity: TP/(TP + FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

## Prediction

We can use our PCA model for predictions of new-unseen cases:

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")

```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize Patient 2 based on the results.