---
title: "Class06: R Functions"
author: "Andrew Duong ( PID: A18129684)"
format: pdf
toc: true
---

## Background

Functions are at the heart of using R. Everything we do involves calling and using functions (from data input, analysis to results output).

All functions in R have at least 3 things:

1. A **name** the thing we use to call the function. 
2. One or more input **arguments** that are comma separated
3. The **body**, lines of code between curly brackets { } that does the work of the function.

## A first function

Let's write a silly wee function to add some numbers:

```{r}
add <- function(x) {
  x + 1
}
```

Let's try it out
```{r}
add(100)
```

Will this work

```{r}
add( c(100,200,300) )
```
Modify to be more useful and add more than just 1

```{r}
add <- function(x, y=1) {
  x + y
}
```


```{r}
add(100, 10)
```
Will this work?
```{r}
add(100)
```

```{r}
plot(1:10, col="blue", typ="o")
```

```{r}
log(10, base = 10)
```

> **N.B.** Input arguments can be either **required** or **optional**. The later have a fall-back default that is specified in the function code with an equals sign. 


```{r}
#add(x=100,y=200,z=300)
```


## A second function

All functions in R look like this

```{r}
name <- function(arg) {
  body
}
```

The `sample()` function in R ...

```{r}
sample(1:10, size = 4)
```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, size = 12, replace = "T")
```

> Q. Write the code to generate a random 12 nucleotide long DNA sequence 

```{r}
bases <- c("A", "G", "T", "C")
sample(bases, size = 12, replace = TRUE)
```

Q. Write a first version function called `generate_dna()` that generates a user specified length `n` random DNA sequence?

```{r}
generate.dna <- function(n) {
  bases <- c("A", "G", "T", "C")
  dna <- sample(bases, size = n, replace = TRUE)
}
```
```{r}
generate.dna(12)
```

> Modify your function to return a FASTA like sequence so rather than [1] "G" "C" "A" "A" "T" we want "GCAAT"

```{r}
generate.dna <- function(n) {
  bases <- c("A", "G", "T", "C")
  dna <- sample(bases, size = n, replace = TRUE)
  return(paste(dna, collapse=""))
}
```
```{r}
generate.dna(12)
```

> Q. Give the user an option to return FASTA format outpit sequence or standard multi-element vector format 

```{r}
generate_dna <- function(n=6, fasta=TRUE) {
  bases <- c("A", "G", "T", "C")
  ans <- sample(bases, size = n, replace = TRUE)
  
  if(fasta) {
    ans <- paste(ans, collapse="")
    cat("Hello...")
  } else  {
  cat("...is it me you are looking for...")
}

return(ans)
}
```

```{r}
generate_dna(10)
generate_dna(10, fasta=F)
```

## A new cool function

> Q. Write a function called `generate_protein()` that generates a user specified length protein sequence in FASTA like format?

```{r}
generate_protein <- function(n, fasta = TRUE) {
  AA <- c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
  ans <- sample(AA, size=n, replace = T)
  ans <- paste(ans, collapse="")
  return(ans)
}
```

```{r}
generate_protein(7)
```


> Q. Use your new `generate_protein()` function to generate sequences between lgenth 6 and 12 amino-acids in length and check if any of these are unique in nature (i.e found in the NR database at NCBI)?

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
```

Or we could do a `for()` loop:

```{r}
for(i in 6:12) {
  cat(">", i, sep="", "\n")
  cat(generate_protein(i), "\n")
}
```



